---
title: "Project 1 - Chess Tournament Data"
author: "Stefano Biguzzi"
date: "9/9/2020"
output:
  html_document:
    includes:
      in_header: header.html
    # css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

# Introduction
In this project, you’re given a text file with chess tournament results where the information has some structure.

Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:

* Player’s Name
* Player’s State
* Total Number of Points
* Player’s Pre-Rating
* Average Pre Chess Rating of Opponents

# Loading packages
These are the packages that will be necessary to complete the following conversion of the Chess Tournament Data.
```{r loading-packages, comment=NA, warning=FALSE, echo=TRUE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

# Load raw data from Github
The first step is to load the data into r from the text file.
```{r load-raw-data, echo=TRUE}
setwd("C:/Users/biguz/Desktop/CUNY Data Science/Fall2020/Data 607/Projects/Project 1")
tournament_data <- readLines("tournamentinfo.txt", warn=FALSE)
```

This data gets read into r as a one large string so we will need to use some string operations to get it to a usable data frame and get it ready to export the data into a csv file.

```{r example-df, results='asis'}
kable(head(data.frame(tournament_data),7), caption = "How initial load of data looks: (first 7 rows as example)") %>%
  kable_minimal()
```

We want to change the data so the lines are gone and all the data between the lines shows up in one row and not two.

# Data munging to create readable dataframe
In this section I used some techniques to change the data from the table shown above to a more readable data frame, where it would be possible to create the CSV file requested.

1. I removed the lines from the data string and replaced where there was more than one space between characters.
```{r cleaning-string, echo=T}
lines <-
  c('-----------------------------------------------------------------------------------------')
tournament_data <- tournament_data[!(tournament_data == lines)]
tournament_data <- str_replace(tournament_data," {2}","")
```

2. I created a data frame. Creating this data frame helps me in a later step by allowing me to merge odd rows with even rows. I also noticed an extra column was created so I removed it as it was duplicate information.
```{r creating-dataframe, echo=T}
#Create dataframe
tournament_raw_df <- data.frame(do.call(rbind, strsplit(tournament_data, "|", fixed=TRUE)))
tournament_raw_df <- tournament_raw_df[-c(11)]
```

3. I realized that for each participant of the tournament there are 2 rows. I had to merge the even row to each odd row using "|" as a separator. After merging the rows I removed every odd row since it was now excess data.
```{r merge-row-data, echo=TRUE, warning=FALSE, comment=NA}
#Merge row data
for (column in names(tournament_raw_df)){
  for (i in 1:nrow(tournament_raw_df)){
    if((i %% 2) > 0){
		  tournament_raw_df[i,column] <-
		    paste(tournament_raw_df[i,column], tournament_raw_df[i+1,column],sep="|")}}}

#Removing odd rows
toKeep <- seq(3,nrow(tournament_raw_df),2)
tournament_raw_df <- tournament_raw_df[toKeep,]
```

4. I rejoined the data frame into a new data frame splitting on the separator "|". This allowed me to create a more readable data frame.
```{r joining-everything-together, echo=TRUE, warning=FALSE, comment=NA}
#Creating new clean dataframe
df_args <- c(tournament_raw_df, sep="|")
tournament_raw_df <- do.call(paste,df_args)
tournament_df <- data.frame(do.call(rbind, strsplit(tournament_raw_df, "|", fixed=TRUE)))

#Renaming columns
tournament_df <- rename(tournament_df,
  c("Pair Num" = "X1", "State" = "X2", "Player Name" = "X3", "USCF ID / Rtg (Pre->Post)" = "X4",
    "Total Pts" = "X5","N Count" = "X6","Round 1 Outcome" = "X7", "Round 1 Pieces" = "X8", 
    "Round 2 Outcome" = "X9", "Round 2 Pieces" = "X10", "Round 3 Outcome" = "X11",
    "Round 3 Pieces" = "X12", "Round 4 Outcome" = "X13", "Round 4 Pieces" = "X14",
    "Round 5 Outcome" = "X15", "Round 5 Pieces" = "X16", "Round 6 Outcome" = "X17",
    "Round 6 Pieces" = "X18", "Round 7 Outcome" = "X19", "Round 7 Pieces" = "X20"))
```

5. I renamed the values for which the chess pieces were identified for each round. So "W" became "White" and "B" became "Black"
```{r replace-pieces, echo=TRUE}
for (col in names(tournament_df)){
  if (grepl("Pieces",col)){
    tournament_df[,c(col)] = ifelse(grepl("W",tournament_df[,c(col)]),"White","Black")}}
```

6. I decided to continue splitting the data frame to create individual columns for each participant USCF ID and each participants rating pre tournament and post tournament rating
```{r splitting-id-rating, echo=TRUE}
#Creating ID column
tournament_df$USCFID <-
  unlist(lapply(
    strsplit(as.character(tournament_df[,c("USCF ID / Rtg (Pre->Post)")]), " / "), '[', 1))

#Creating pre-post column
tournament_df$"Rtg (Pre->Post)" <-
  unlist(lapply(
    strsplit(as.character(tournament_df[,c("USCF ID / Rtg (Pre->Post)")]), " / "), '[', 2))

#Removing old id/rating column
tournament_df <- tournament_df[, -which(names(tournament_df) %in% "USCF ID / Rtg (Pre->Post)")]

#Creating pre rating column
tournament_df$PreRating <-
  unlist(lapply(
    strsplit(as.character(tournament_df[,c("Rtg (Pre->Post)")]), "->"), '[', 1))

#Creating post rating column
tournament_df$PostRating <-
  unlist(lapply(
    strsplit(as.character(tournament_df[,c("Rtg (Pre->Post)")]), "->"), '[', 2))

#Removing old pre/post rating column
tournament_df <- tournament_df[,-which(names(tournament_df) %in% "Rtg (Pre->Post)")]
```

7. I cleaned up the new pre and post rating columns by removing any excess string. (Some of the ratings had a P followed by a list of numbers, which I removed as it was not relevant to this project.)
```{r clean-up-ratings, echo=TRUE}
#Remove "R:" from pre rating number
tournament_df$PreRating <- gsub("R: ", "", tournament_df$PreRating)

#Remove P from pre and post rating
tournament_df$PreRating <- gsub("()P.*", "", tournament_df$PreRating)
tournament_df$PostRating <- gsub("()P.*", "", tournament_df$PostRating)
```

8. I split the round outcome column into round outcome and round opponent columns. This way I could easily match a round opponent to his pre tournament ranking.
```{r create-round opponent, echo=TRUE}
#Create new columns and populate values
for (col in names(tournament_df)){
  if (grepl("Outcome",col)){
    #Opponent column
    tournament_df[,c(gsub("Outcome", "Opponent", col))] <-
      unlist(lapply(strsplit(as.character(tournament_df[,c(col)]), "  "), '[', 2))
    #Outcome column
    tournament_df[,c(col)] <- gsub("([A-Z]).*","\\1",tournament_df[,c(col)])}}

#Clean up white spaces
for (i in names(tournament_df)) {
  tournament_df[[i]] <-
    trimws(tournament_df[[i]], which = c("both", "left", "right"), whitespace = "[ \t\r\n]")}
```

9. I created the opponent pre tournament chess rating by matching the round opponent pair number to the tournament data frame pair number, grabbing the value in the *PreRating* column, and writing it in the corresponding row and column
```{r create-opp-rtg, echo=TRUE}
for (i in seq(nrow(tournament_df))){
  for (col in names(tournament_df)){
    if (grepl("Opponent",col)){
        tournament_df[i,c(gsub("Opponent","Opp PreRtg",col))] <-
          ifelse(
            is.na(tournament_df[i,col]), NA,
            tournament_df$PreRating[tournament_df$`Pair Num` == tournament_df[i,col]])}}}
```

10. Finally, I reorganized the columns for readability. This was the final raw data frame. This is still a raw dataframe as it is too wide making it very unreadable, as seen in the example below.
```{r reorganize-tournament-df}
tournament_df <- tournament_df[,c("Pair Num", "State", "Player Name", "USCFID", "PreRating",
                                  "PostRating", "Total Pts", "N Count", "Round 1 Opponent",
                                  "Round 1 Opp PreRtg", "Round 1 Outcome", "Round 1 Pieces",
                                  "Round 2 Opponent", "Round 2 Opp PreRtg", "Round 2 Outcome",
                                  "Round 2 Pieces", "Round 3 Opponent", "Round 3 Opp PreRtg",
                                  "Round 3 Outcome", "Round 3 Pieces", "Round 4 Opponent",
                                  "Round 4 Opp PreRtg", "Round 4 Outcome", "Round 4 Pieces",
                                  "Round 5 Opponent", "Round 5 Opp PreRtg", "Round 5 Outcome",
                                  "Round 5 Pieces", "Round 6 Opponent", "Round 6 Opp PreRtg",
                                  "Round 6 Outcome", "Round 6 Pieces", "Round 7 Opponent",
                                  "Round 7 Opp PreRtg", "Round 7 Outcome", "Round 7 Pieces")]
```
```{r}
kable(head(tournament_df,5),caption = "Data dump of chess tournament (example first 5 rows)") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = F)
```

# Creating the CSV file
For this project we were asked to generate a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:

* Player’s Name
* Player’s State
* Total Number of Points
* Player’s Pre-Rating
* Average Pre Chess Rating of Opponents

The first 4 columns of the csv file were completed in the previous section. I used this section to create the final column for the CSV table *Average Pre Chess Rating of Opponents*.

1. I converted the opponent pre rating fields, which I summed in the next step, into numeric columns using a for loop.
```{r converting-numeric, echo=TRUE}
for (col in names(tournament_df)){
  if (grepl("PreRtg",col)){
    tournament_df[,c(col)] <- as.numeric(tournament_df[,c(col)])}}
```

2. I summed up all the pre game ratings of each opponent while also summing up the number of games played. I did this by using the *rowSums* function while ignoring the NA since some participants did not play a full 7 rounds.
```{r summing-prertgs, echo=TRUE}
#Creating pre rating sum
tournament_df$OppPreRtgSum <-
  rowSums(tournament_df[,c("Round 1 Opp PreRtg", "Round 2 Opp PreRtg", "Round 3 Opp PreRtg",
                           "Round 4 Opp PreRtg", "Round 5 Opp PreRtg", "Round 6 Opp PreRtg",
                           "Round 7 Opp PreRtg")], na.rm = TRUE)

#Creating games played
tournament_df$GamesPlayed <-
  rowSums(!is.na(tournament_df[,c("Round 1 Opp PreRtg", "Round 2 Opp PreRtg", "Round 3 Opp PreRtg",
                                  "Round 4 Opp PreRtg", "Round 5 Opp PreRtg", "Round 6 Opp PreRtg",
                                  "Round 7 Opp PreRtg")]))
```

3. I divided the total sum of pre tournament ratings by the number of games played to get the average opponent rating for each tournament participant.
```{r creating-avg-opp-rtg, echo=TRUE}
tournament_df <- tournament_df %>%
  mutate(AvgOppRtg = OppPreRtgSum/GamesPlayed)
```

4. I created the export csv table as asked in number 5. As seen below:
```{r renaming-columns}
tournament_df <- rename(tournament_df, c("Player's Name" = "Player Name",
                                         "Player's State" = "State",
                                         "Total Number of Points" = "Total Pts",
                                         "Player's Pre-Rating" = "PreRating",
                                         "Average Pre Chess Rating of Opponents" = "AvgOppRtg"))
```
```{r creating-csv}
write.csv(tournament_df[,c("Player's Name", "Player's State","Total Number of Points",
                           "Player's Pre-Rating",
                           "Average Pre Chess Rating of Opponents")],
          "./tournament_data.csv", row.names = FALSE)
```
```{r results='asis'}
tournament_df[,c("Player's Name","Player's State","Total Number of Points",
                 "Player's Pre-Rating","Average Pre Chess Rating of Opponents")] %>%
  kbl(caption = "Final Chess Tournament Table") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
\newpage


***



<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>

